---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data   
### 1. Loading data  
```{r}  
zipFileName <- "activity.zip"    
csvFileName <- "activity.csv"  
activity <- read.csv(unz(zipFileName, csvFileName), colClasses=c("numeric", "character","numeric")) 
head(activity)
```
### 2. Process/transform the data  
```{r} 
activity$date <- as.Date(activity$date, "%Y-%m-%d")  
good <- complete.cases(activity$steps)   
activity_complete <- activity[good,]   
```
## What is mean total number of steps taken per day?  
### 1. Calculate the total number of steps taken per day  
```{r}  
good <- complete.cases(activity$steps)  
activity_complete <- activity[good,]
library(dplyr)
steps_per_day <- activity_complete %>% group_by(date) %>% summarize(steps = sum(steps))
head(steps_per_day)
```
### 2. Make a histogram of the total number of steps taken each day  
```{r} 
hist(
        steps_per_day$steps, 
        breaks=10,  
        main="Total Number of Steps Per Day",  
        xlab="Number of Steps" ,  
        ylab="Frequency" 
        )  
```
### 3. Calculate and report the mean and median of the total number of steps taken per day  
```{r}
steps_mean <- as.character(round(mean(steps_per_day$steps),2))
steps_median <- as.character(round(median(steps_per_day$steps),2))
```
The mean of the total number of steps taken per day is `r steps_mean`   
The median of the total number of steps taken per day is `r steps_median`  
## What is the average daily activity pattern?
### 1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)  
```{r}
steps_per_interval <- activity_complete %>% group_by(interval) %>% summarize(steps = mean(steps))
head(steps_per_interval)
plot(
        steps_per_interval$interval, 
        steps_per_interval$steps,  
        type="l",
        main="Average Number of Steps Taken of 5-minute Interval",  
        xlab="5-minute Interval" ,  
        ylab="Average Number of Steps Taken" 
        ) 
```

### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?  
```{r}
steps_max_interval <- steps_per_interval[steps_per_interval$steps==max(steps_per_interval$steps),1]
```
The 5-minute interval that contains the maximum number of steps is `r steps_max_interval`  
## Imputing missing values
### 1. Calculate and report the total number of missing values in the dataset.  
```{r}
na_total <- as.character(nrow(activity) - nrow(activity_complete))
```
The total number of missing values in the dataset is `r na_total`  
### 2. Devise a strategy for filling in all of the missing values in the dataset.    
Use the mean for that 5-minute interval to fill in all of the missing values.  
### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.  
```{r}
activity_na_filled <- data.frame(activity)
index <- as.vector(row(activity_na_filled))
for (i in index) {
        if (is.na(activity_na_filled[i,"steps"])) {
                activity_na_filled[i,"steps"] <- steps_per_interval[steps_per_interval$interval==activity_na_filled[i,"interval"],"steps"]
        }
}
head(activity_na_filled)
```
### 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r} 
steps_per_day_na_filled <- activity_na_filled %>% group_by(date) %>% summarize(steps = sum(steps))
hist(
        steps_per_day_na_filled$steps, 
        breaks=10,  
        main="Total Number of Steps Per Day (NA filled with 5-minute interval mean)",  
        xlab="Number of Steps" ,  
        ylab="Frequency" 
        )  
steps_mean_na_filled <- as.character(round(mean(steps_per_day_na_filled$steps),2))
steps_median_na_filled <- as.character(round(median(steps_per_day_na_filled$steps),2))
```
The mean of the total number of steps taken per day is `r steps_mean_na_filled`   
The median of the total number of steps taken per day is `r steps_median_na_filled`   
These values differ slightly from the first part of the assignment and has little impact on the estimates since the 5-minute interval mean is used to fill in the missing values.   
## Are there differences in activity patterns between weekdays and weekends?  
### 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.  
```{r} 
isWeekday <- function(d){
        if (weekdays(d) %in% c('Saturday', 'Sunday')) {
                as.factor("weekend")
        } else {
                as.factor("weekday")
        }
}
activity_na_filled$is_weekday = sapply(activity_na_filled$date,isWeekday)
head(activity_na_filled)
```
### 2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.  

```{r}
steps_per_interval_na_filled <- activity_na_filled %>% group_by(is_weekday, interval) %>% summarize(steps = mean(steps))
library(lattice)
xyplot(
        steps~interval | is_weekday, data = steps_per_interval_na_filled,
        layout = c(1,2),
        type = 'l',
        xlab = 'Interval',
        ylab = 'Number of steps'
        )
```  
